---
title: "Data Manipulation in R with dplyr"
author: "GT"
date: "November 17, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction to dplyr and tbls

Introduction to the dplyr package and the tbl class. Learn the
philosophy that guides dplyr, discover some useful applications
of the dplyr package, and meet the data structures that dplyr 
uses behind the scenes.

#### Load the dplyr and hflights package

Welcome to the interactive exercises part of your `dplyr` course.
Here you will learn the ins and outs of working with `dplyr`. 
`dplyr` is an R package, a collection of functions and data sets
that enhance the R language.

Throughout this course you will use `dplyr` to analye a data set
of airline flight data containing fligths that departed from 
Houston. This data is stored in a package calle `hflights`.

```{r}
# Load the dplyr package
library(dplyr)

# Load the hflights package
library(hflights)

# Call both head() and summary() on hflights
head(hflights)
summary(hflights)
```

#### Explore the data set

A data scientist must be familiar with his or her data. Experiment
with the data set in the console and maybe try to generate
some insightful plots.

How many observations and how many variables are contained in
the hflight data set ?

```{r}
dim(hflights)
```

The variables are stored in the columns, whereas the observations 
are stored in the rows of this data set. As you can see, this
data set is relatively large. In fact, it's about 25Mb in size.
For your convenience, we will continue the course's exercises with 
a random subset of the dataset.

```{r}
set.seed(10)
hflights <- hflights[sample(nrow(hflights),0.1*nrow(hflights)),]
dim(hflights)
```


#### Convert data.frame to table

As Garrett explained, a **tbl** is just a special kind of data.frame.
They make your data easier to look at, but also easier to work with.
On top of this, it is straightforward to derive a tbl from a data.frame
structure using `tbl_df()`.

The tbl format changes how R displays your data, but it does not
change the data's underlying data structure. A tbl inherits the
original class of its input, in this case, a data.frame. This
means that you can still manipulate the tbl as if it were a data.frame.
In other words, you can do anything with the `hflights` tbl that you 
could do with the `hflights` data.frame.

```{r}
# Convert the hflights data.frame into a hflights tbl
hflights <- tbl_df(hflights)

# Display the hflights tbl
hflights

# Create the object carriers
carriers <- hflights$UniqueCarrier

class(hflights)
```

Note that the resulting `hflights` tbl is also a data.frame; you
can assert this by executing `class(hflights)` in the console. 
This means you can do everything with a tbl that you could do with
a data.frame.

#### Changing labels of hflights

A bit of cleaning would be a good idea since the `UniqueCarrier`
variable of `hflights` uses a confusing code system.

To do this, let's work with a lookup table, that comes in the
form of a named vector. When you subset the lookup table with
a character string, R will return the values of the lookup table
that correspond to the names in the character string. To see
fow this works, run following code:

```{r}
two <- c("AA", "AS")
lut <- c("AA" = "American",
         "AS" = "Alaska",
         "B6" = "JetBlue")
two <- lut[two]
two
```

```{r}
# Both the dplyr and hflights packages are loaded into workspace
lut <- c("AA" = "American", "AS" = "Alaska", "B6" = "JetBlue", "CO" = "Continental", 
         "DL" = "Delta", "OO" = "SkyWest", "UA" = "United", "US" = "US_Airways", 
         "WN" = "Southwest", "EV" = "Atlantic_Southeast", "F9" = "Frontier", 
         "FL" = "AirTran", "MQ" = "American_Eagle", "XE" = "ExpressJet", "YV" = "Mesa")

# Add the Carrier column to hflights
hflights$Carrier <- lut[hflights$UniqueCarrier]

# Glimpse at hflights
glimpse(hflights)
```

Let's try a similar thing, but this time to change the labels
in the `CancellationCode` column. This column lists reasons why
a flight was cancelled using a non-informative alphabetical code.

```{r}
unique(hflights$CancellationCode)
```

A lookup table `lut` has already been created for you, that
converts the alphabetical codes into more meaningful strings.

```{r}
# The lookup table
lut <- c("A" = "carrier", "B" = "weather", "C" = "FFA", "D" = "security", "E" = "not cancelled")

# Add the Code column
hflights$Code <- lut[hflights$CancellationCode]

# Glimpse at hflights
glimpse(hflights)
```

### Select and mutate

Get familiar with dplyr's manipulation verbs. Meet the five
verbs and then practice using the mutate and select verbs.

#### The five verbs and their meaning

The `dplyr` package contains five key data manipulation functions,
also called verbs:

- `select()`, which returns a subset of the columns
- `filter()`, that is able to return a subset of the rows
- `arrange()`, that reorders the rows according to single or
  multiple variable
- `mutate()`, used to add columns from existing data
- `summarise()`, which reduces each group to a single row
  by calculating aggregate measures

What order of operations should we use to find the average value
of the `ArrDelay` variable for all American Airline flights
in the `hflights` tbl ?

```{r}
filter(hflights, hflights$UniqueCarrier == "AA") %>% summarise(mean=mean(ArrDelay, na.rm=TRUE))
```

#### Choosing is not losing! The select verb

To answer the simple question wheter flight delays tend to shrink
or grow during a flight, we can safely discard a lot of the variables
of each flight. To select only the ones that matter, we can use
`select()`.

As an example, take the following call, that selects the variables
`var1` and `var2` from the data frame `df`.

```
select(df, var1, var2)
```

You can also use `:` to select a range of variables and `-` to
exclude some variables, similar to indexing a data.frame with 
square brackets. You can use both variable's names as well as 
integer indexes. This call selects the four first variables
except for the second one of a data frame `df`:

```
select(df, 1:4, -2)
```

`select()` does not change the data frame it is called on;
you have to explicitly assign the result of `select()` to a 
varible to store the result.

```{r}
# Print out a tbl with the four columns of hflights related to delay
select(hflights, ActualElapsedTime, AirTime, ArrDelay, DepDelay)

# Print out the columns Origin up to Cancelled of hflights
select(hflights, Origin:Cancelled)

# Answer to last question: be concise!
select(hflights, Year:DayOfWeek, ArrDelay:Diverted)
```

Bear in mind that `select()` does not modify the original dataset.
Instead, `select()` returns a new, modified copy. Therefore,
choosing is not losing! You should save the result to a new variable
if you want to be able to use it later.

#### Helper functions for variable selection

`dplyr` comes with a set of helper functions that can help you
select groups of variables inside a `select()` call:

- `starts_with("X")`: every name that starts with `"X"`
- `end_with("X")`: every name that ends with `"X"`
- `contains("X")`: every name that contains `"X"`
- `matches("X")`: every name that matches `"X"`, where `"X"` can be
  a regular expression
- `num_range("x", 1:5)`: the variables names `x01`, `x02`, `x03`,
  `x04`, `x05`
- `one_of(x)`: every name that appears in `x`, which should be 
  a character vector

Pay attention here: when you refer to columns directly inside `select()`,
you don't use quotes. If you use the helper functions, you do use quotes.

```{r}
# Print out a tbl containing just ArrDelay and DepDelay
select(hflights, ends_with("Delay"))

# Print out a tbl as described in the second instruction, using both helper functions and variable names
select(hflights, starts_with("Cancell"), ends_with("Num"), UniqueCarrier)

# Print out a tbl as described in the third instruction, using only helper functions.
select(hflights, ends_with("Time"), ends_with("Delay"))
```

This exercise demonstrates the power of the helper functions. 
They can limit the amount of coding drastically. 
Beware of too obscure queries however; you still want your code to be
clear for fellow data scientists!

#### Comparison to basic R

To see the added value of the `dplyr` package, it is useful to 
compare its syntax with basic R. Up to now, you have only considered
functionality that is also available without the use of `dplyr`.
The elegance and ease-of-use of `dplyr` is a great plus though.

```{r}
# Finish select call so that ex1d matches ex1r
ex1r <- hflights[c("TaxiIn", "TaxiOut", "Distance")]
ex1d <- select(hflights, Distance:TaxiOut)

# Finish select call so that ex2d matches ex2r
ex2r <- hflights[c("Year", "Month", "DayOfWeek", "DepTime", "ArrTime")]
ex2d <- select(hflights, Year:ArrTime, -DayofMonth)

# Finish select call so that ex3d matches ex3r
ex3r <- hflights[c("TailNum", "TaxiIn", "TaxiOut")]
ex3d <- select(hflights, starts_with("T"))
```

As you can see, the `dplyr` syntax can simplify things a lot. 
The key to success is to look for structure in your column names. 
Time to explore another data manipulation verb!

#### Mutating is creating

#### Add multiple variables using mutate

#### Recap on mutate and select



