---
title: "Titanic"
author: "GT"
date: "3 mai 2016"
output: html_document
---

## Titanic: Machine Learning from Disaster

### Datacamp tutorial

When the Titanic sank, 1502 of the 2224 passengers and crew got killed. One of the main reasons for 
this high level of casualties was the lack of lifeboats on this supposedly unsinkable ship.

Those that have seen the movie know that some individuals were more likely to survive the sinking 
(lucky Rose) than others (poor Jack). In this course you wil apply machine learning techniques to 
predict a passenger's chance of surviving using R.

Let's start with loading in the training and testing set into your R environment. You will use the 
training set to build your model, and the test set to validate it. The data is stored on the web as 
CSV files. You can load this data with the `read.csv()` function.

```{r}
train = read.csv("train.csv") # http://s3.amazonaws.com/assets.datacamp.com/course/Kaggle/train.csv
test = read.csv("test.csv") # http://s3.amazonaws.com/assets.datacamp.com/course/Kaggle/test.csv
```

Before starting with the actual analysis, it's important to understand the structure of your data. 
Both `test` and `train` are data frames, R's way of representing a dataset. You can easily explore a 
data frame using the `str()` function. `str()` gives you information such as the data types in the 
data frame (e.g. `int` for integer), the number of observations, and the number of variables.

```{r}
str(train)
str(test)
```

How many people in your training set survived the disaster with the Titanic? To see this, you can use 
the `table()` command in combination with the `$`-operator to select a single column of a data frame:

```{r}
# absolute numbers
table(train$Survived) 

# proportions
prop.table(table(train$Survived))
```

You see that 549 individuals died (62%) and 342 survived (38%). A simple prediction heuristic could thus 
be "majority wins": you predict every unseen observation to not survive.

In general, the `table()` command can help you to explore which variables have predictive value. 
For example, maybe gender could play a role as well? For a two-way comparison, also including gender, 
you can use

```{r}
table(train$Sex, train$Survived)
```

To get proportions, you can again wrap `prop.table()` around `table()`, but you'll have to specify whether 
you want row-wise or column-wise proportions: This is done by setting the second argument of `prop.table()`, called `margin`, to 1 or 2, respectively.

```{r}
prop.table(table(train$Sex, train$Survived), margin = 1)
```

Female had over a 50% chance of surviving.

Another variable that could influence survival is age: it's probable children were saved first. You can test 
this by creating a new column with a categorical variable `child`.

To add this new variable you need to do two things:

1. Create a new column `Child`, which is done through the `$` operator.
  
```{r}
train$Child <- NA
```
  
2. Provide the values for each observation (i.e., row) based on the age of the passenger. You can use a 
boolean test inside square brackets for this. Set the `Child` variable to `1` for passengers aged less
than 18 years old, and `0` the others.

```{r}
train$Child[train$Age < 18] <- 1
train$Child[train$Age >= 18] <- 0
```
  
3. Do a two-way comparison on the number of children vs adults that survived, in row-wise proportions

```{r}
prop.table(table(train$Child, train$Survived), margin = 1)
```

While less obviously than gender, Age also seems to have an impact on survival.

In your training set, females had over a 50% chance of surviving and males had less than a 50% 
chance of surviving. Hence, you could use this information for your first prediction: all females 
in the test set survive and all males in the test set die.

You use your test set for validating your predictions. You might have seen that, contrary to 
the training set, the test set has no `Survived` column. You add such a column using your 
predicted values. Next, when uploading your results, Kaggle will use this column (= your predictions) 
to score your performance.

```{r}
# Copy of test
test_one <- test

# Initialize a Survived column to 0
test_one$Survived = 0

# Set Survived to 1 if Sex equals "female"
test_one$Survived[test_one$Sex == "female"] = 1
```

